---
title: "fenosoa_2025_manuscript_code"
author: "Kacie Ring"
date: "2025-01-03"
output: 
  html_document: 
    toc: true
    toc_float: true
---

The code below was used for the statistical analyses and data visulaization for the manuscript titled, "Gut passage in lemurs enhances the germination of mistletoe seeds Bakerella gonoclada, in a Madagascar rainforest" 

Authors: Zo Samuel Ella Fenosoa, Cara E. Brook, Amy E. Dunham, Hajanirina F. Rakotomanana, Hanta Razafindraibe, Kacie A. Ring



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


# Load the packages
library(survival)
library(survminer)
library(ggsignif)
library(ggplot2)
library(tidyverse)
library(ggsignif)
library(lubridate)
library(survival)
library(boot)
library(car)
library(dplyr)
library(nlme)
library(MASS)
library(lme4)

```



#### Data cleaning, replacing NAs with zero values 

```{r}

#read in the dataframe 
germ <- read.csv("germination.csv")


#datacleaning
germ <- germ %>% 
  mutate(hypocot_lgth_final = replace_na(hypocot_lgth_final, 0),  #changing NA values to 0
         time_to_germ = replace_na(time_to_germ, 0),              #changing NA values to 0 
         leaves_lgth_final = replace_na(leaves_lgth_final, 0),     #changing NA values to 0 
         death = 1,                                                #changing NA values to 0
         seed_type = as.factor(seed_type),
         source_seed = as.factor(source_seed))

```


### Statistical GLMM Model 1: germination probability model 

```{r statstical model 1}

#Model 1: Probability of germination, binomial distribution 



#we want to set hand-pulled seeds as the reference for the model so we can more easily compare how defecated changes germination
germ$seed_type <- relevel(germ$seed_type, "plant") #set plant as reference for the model 
m1 <- glmer(germ_Y_N ~ seed_type + (1|source_seed), family = "binomial", data = germ)
summary(m1) #model output 
ranef(m1) #intercepts of the random effect 
      
#fitting the full model without a random effect to see if its important/significant 
reduced_model <- glm(germ_Y_N ~ seed_type, family = binomial, data = germ)
summary(reduced_model)

#Chi-sq for model comparison 
model_compare <- anova(m1, reduced_model, test = "Chisq")
summary(model_compare)

#create the table for the supplemental information 
#tab_model(m1,  pred.labels = c("Intercept", "Seed source (Defecated)"), 
         # dv.labels = "Germinated (Y/N)")

```



### Figure 1a plot: Percent germination  


```{r data cleaning and plotting for fig 1, warnings=FALSE, fig.show="hold", echo=FALSE, output=FALSE}

#data cleaning and table creation for figure 
germ_summary <- germ %>%
  group_by(seed_type) %>%
  summarise(
    n_germinated = sum(germ_Y_N == 1),
    total = n(),
    percentage = (n_germinated/total) * 100
  )

germ_summary
```


```{r}
#reorder the levles of the factor for aesthetic reasons 

germ_summary$seed_type <- factor(germ_summary$seed_type, levels = c("plant", "feces"))

# Create the plot
Figure_1a <- ggplot(germ_summary, aes(x = seed_type, y = percentage, fill = seed_type)) + # Added fill aesthetic
  geom_bar(stat = "identity", width = 0.6) +
  geom_signif(
    comparisons = list(c("plant", "feces")),
    map_signif_level = TRUE,
    annotations = "***",
    textsize = 6,
    vjust = 0.6
  ) +
  scale_x_discrete(labels = c("Hand-removed\n(N = 48)", "Defecated\n(N = 18)")) +
  scale_fill_manual(values = c("darkolivegreen", "chocolate4"), # Define colors
                    labels = c("Hand-removed", "Defecated"),     # Define labels
                    name = "Seed source") +                      # Legend title
  labs(
    x = "Seed source",
    y = "Germination percentage (%)"
  ) +
  ylim(0, 105) +
  theme_bw() +
  theme(
    axis.text = element_text(size = 9.5),
    axis.title = element_text(size = 11),
    legend.position = "bottom"                                  # Position legend at bottom
  )

#save the figure 
ggsave("Figure_1a.png", plot = Figure_1a, width = 10, height = 7, device = png, dpi = 500)

Figure_1a

```


### Time to germination stats, exploratory data analysis 


```{r}

#basis stats for time to germination 

germ_stats <- germ %>%
  filter(time_to_germ > 0) %>%
  group_by(seed_type) %>% 
  summarize(Min = min(time_to_germ),
             Q1 = quantile(time_to_germ, .25),
            mean = mean(time_to_germ), 
             Q3 = quantile(time_to_germ, .75),
             Max = max(time_to_germ),
            sd = sd(time_to_germ)
             )

germ_stats
```



### Figure 1b plot:  A cumulative incidence plot visually represents the probability of a specific event occurring within a given 


```{r}

# Remove ggsurvfit from your package loading section

# For Figure 1b, replace the ggsurvfit code with this:

# Create the survival fit
fit_germ <- survfit(Surv(time_to_germ, germ_Y_N) ~ seed_type, data=germ)

# For germination plot (Figure 1b)

Figure_1b <- ggsurvplot(
  fit_germ,
  data = germ,
  fun = "event",  # For cumulative incidence
  size = 1,
  conf.int = TRUE,
  palette = c("chocolate4", "darkolivegreen"),
  legend.labs = c("Defecated", "Hand-removed"),
  legend.title = "Seed source",
  legend = "bottom",
  xlab = "Time (days)",
  ylab = "Cumulative proportion of seeds germinated (%)",
  xlim = c(0, 7),
  ggtheme = theme_bw()
)
# Extract the actual ggplot object
Figure_1b <- Figure_1b$plot


#save the plot 
ggsave("Figure_1b.pdf", plot = Figure_1b, width = 10, height = 7)

Figure_1b

```




### Figure 1c and Survival analysis ( time-to-event data anlysis)

Survival data, also called time-to-event data, consist of a distinct start time and end time and analysis aims to predict the time until an event occurs 

```{r Kaplan-Meier survival curve}
# Analysis for survival (Figure 1c)
fit_surv <- survfit(Surv(time_to_death, death) ~ seed_type, data=germ)

# Statistical tests for survival

surv_diff <- survdiff(Surv(time_to_death, death) ~ seed_type, data=germ)
cox_model <- coxph(Surv(time_to_death, death) ~ seed_type, data=germ)


# For survival plot (Figure 1c)

Figure_1c <- ggsurvplot(
  fit_surv,
  data = germ,
  size = 1,
  palette = c("darkolivegreen", "chocolate4"),
  conf.int = TRUE,
  pval = TRUE,
  pval.coord = c(0, 0.1),
  risk.table = FALSE,
  legend = "bottom",
  xlab = "Time (days)",
  legend.labs = c("Hand-removed Seeds", "Defecated Seeds"),
  ggtheme = theme_bw()
)

# Extract the actual ggplot object
Figure_1c <- Figure_1c$plot


# Print statistical results
cat("Survival Analysis Results:\n")
cat("\nLog-rank test:\n")
print(surv_diff)
cat("\nCox proportional hazards model:\n")
print(summary(cox_model))

# Save plots
ggsave("Figure_1b.pdf", plot = Figure_1b, width = 10, height = 7)
ggsave("Figure_1c.pdf", plot = Figure_1c, width = 10, height = 7)


```




### GLMM Model 2 prediciting hypocotyl length 


```{r statistial model 3}

#Model 3: hypocotyl length between seeds sources 

time_to_germ_dat <- germ %>%
  filter(germ_Y_N == 1)

hist(time_to_germ_dat$hypocot_lgth_final) #right skewed data Y variable 
mean(time_to_germ_dat$hypocot_lgth_final)
var(time_to_germ_dat$hypocot_lgth_final)
# mean > variance 

#glm with possion distribution becasue mean > variance 
m4 <- glmer(hypocot_lgth_final ~ seed_type + (1|source_seed), family = poisson(link = "log"), data = germ)
#model results 
summary(m4)

#lets check out the mode where we remove random effects 
reduced_model_2 <- glm(hypocot_lgth_final ~ seed_type, family = poisson(link = "log"), data = germ)

#compare models woth an anova 
anova(m4, reduced_model_2, test = "Chisq")


```


```{r results table}

#tab_model(m4,
        #  pred.labels = c("Intercept", "Seed source (Defecated)"), 
        #  dv.labels = ("Hypocotyl length (cm)"))

```


### Figure 1d plot: Violin plot showing diiferences in hypocotyl length 

```{r Figure 1d}

 Figure_1d <- germ %>% 
 arrange(factor(seed_type, levels = c("feces", "plant"))) %>%
  ggplot(., aes(seed_type, hypocot_lgth_final, fill = seed_type)) +
  geom_violin() +
  geom_jitter(height = 0, width = 0.1) + 
  geom_signif(comparisons = list(c("feces", "plant")), map_signif_level = TRUE,    annotations = "***", textsize = 6) + 
  ylim(0,8) + #adding significance astericks 
  # Increase legend title font size
  scale_fill_manual(values = c("darkolivegreen", "chocolate4"), labels = c("Hand-removed", "Defecated")) +
  labs(y = "Hypocotyl length (mm)", x = "Seed source", color = "Seed source", fill = "Seed source") + 
 scale_x_discrete(labels = c('Hand-removed (N = 48)', 'Defecated (N = 18)')) + 
  theme_bw()

Figure_1d <-Figure_1d + theme(axis.title.y = element_text(size=8), axis.text.y = element_text(size = 10))
Figure_1d <-Figure_1d + theme(legend.position = "bottom") 

ggsave("Figure_1d.png", plot = Figure_1d, width = 10, height = 7, device = png, dpi = 500)

Figure_1d

```


```{r}
source("group_by_summary_stats.R")
group_by_summary_stats(germ, hypocot_lgth_final, seed_type)

```


### Final combined Figure 1 for manuscript

```{r}
library(patchwork)

# Combine the plots with annotations
combined_plot <- (Figure_1a | Figure_1b) / (Figure_1c| Figure_1d)

# Add annotations
combined_plot <- combined_plot + plot_annotation(tag_levels = 'a')

ggsave("my_combined_plot.png", plot = combined_plot, width = 10, height = 8, dpi = 300)

#final figure for panuscript
combined_plot
```


